Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-05 paper/main.tex
Updates

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on their domain knowledge. This preference arises due to several
practical issues with automated algorithms, such as they make obvious errors
and output Markov Equivalence Classes (MECs). To address these challenges, we
propose an iterative method to assist researchers in manually constructing or
improving an existing DAG by combining implied Conditional Independence (CI)
testing with expert knowledge. Our proposed method utilizes a novel measure of
partial association for mixed data to rank violations of implied CIs in a given
DAG, enabling users to prioritize modifications while utilizing their domain
knowledge to decide the orientation of new edges. Empirical results show that
this manual model construction approach achieves performance comparable to
automated algorithms if the user is able correctly orientation edges in one out
of three cases. Additionally, we demonstrate that, in the absence of an expert,
a Large Language Model (LLM) can be potentially determine edge orientations. We
provide the implementation of our approach in both a web tool at: <redacted for
review> and a Python package <redacted for review>.

>>>>

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on domain knowledge. This preference arises due to several practical
challenges with automated algorithms, such as their tendency to make obvious
errors and output Markov Equivalence Classes (MECs) rather than a DAG. To
address these challenges, we propose an iterative method that guides
researchers in manually constructing or improving the fit of an existing DAG.
Our approach combines implied Conditional Independence (CI) testing with a
novel measure of conditional association between variables to rank violations
of implied CIs in a given DAG. This ranking helps prioritize modifications to
the DAG to improve its fit to the data while utilizing expert knowledge to
decide the orientation of new edges. Empirical results show that this guided
manual model construction approach achieves performance comparable to automated
algorithms if we are able to correctly decide the orientation of new edges in
one out of three cases. Additionally, we demonstrate that, in the absence of an
expert, a Large Language Model (LLM) can be potentially used to determine edge
orientations. We provide the implementation of our approach in both a web tool
at: <redacted for review> and a Python package <redacted for review>.
